---
title: "Basic Example"
author: "Ruben Cabrera"
date: "2023-08-01"
output: 
  github_document: default
  html_document: default
# always_allow_html: true
# devtools::install_github("kupietz/kableExtra")
# install.packages("xfun", repos=c("http://rstudio.org/_packages", "http://cran.rstudio.com"))
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic Example

Basic Example

https://www.youtube.com/watch?v=oI1_SV1Rpfc

```{r, message=FALSE, warning=FALSE}

library(nlme)
library(lmerTest)
library(ggplot2)
library(dplyr) # For filter
library(lattice)
# install.packages("sjPlot")
library(sjPlot) # To use plot_model

```

## Data

```{r data}
df <- data.frame(
  Weight=c(102,96,83,79,97,93,79,77,95,87,78,75,93,85,74,72),
  Subjects=c(1,2,3,4,1,2,3,4,1,2,3,4,1,2,3,4),
  Diet=c('A','A','B','B','A','A','B','B','A','A','B','B','A','A','B','B'),
  Weeks=c(0,0,0,0,1,1,1,1,2,2,2,2,3,3,3,3)
)
```


## Plot data

```{r plotdata}
ggplot(df, aes(x = Weeks, y = Weight, group = factor(Subjects), color = factor(Subjects))) + geom_line()+ geom_point() + xlab('Weeks')

```

## Model 1: Weight ~ Weeks+(1|Subjects)

```{r model1}
M1 <- lmer(Weight ~ Weeks+(1|Subjects),REML = F, data = df)
summary(M1)
random.effects(M1)  # ranefe(M1)  random.effects(M1)
```
# Plots Model M1: Weight ~ Weeks+(1|Subjects)

```{r M1_plots}

# Residuals vs Fitted
plot(M1, which = 1)

# Dotplot of random effects
dotplot(ranef(M1, condVar = TRUE))

# Install and load the sjPlot package

# Plot fixed and random effects
sjPlot::plot_model(M1, type = "re")

# Plot general fit

ggplot(df, aes(x = Weeks, y = Weight, color = factor(Subjects))) +
  geom_point() +
  geom_line(aes(group = Subjects), linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "black")

# Extract fitted values
df$fitted <- fitted(M1)

# Plot fitted

ggplot(df, aes(x = Weeks, y = Weight, color = Subjects)) +
  geom_point() +
  geom_line(aes(y = fitted, group = Subjects)) +
  labs(title = "Fitted Values for Each Subject", x = "Weeks", y = "Weight")

```
# Model 2: Weight ~ Weeks+(1+Weeks|Subjects)

```{r model2}
M2 <- lmer(Weight ~ Weeks+(1+Weeks|Subjects),REML = F, data = df)
summary(M2)
random.effects(M2)  # ranefe(M2)

```

# Plots Model M2: Weight ~ Weeks+(1+Weeks|Subjects)

```{r M2_plots}
# Residuals vs Fitted
plot(M2, which = 1)

# Extract fitted values
df$fitted <- fitted(M2)

# Plot fitted

ggplot(df, aes(x = Weeks, y = Weight, color = Subjects)) +
  geom_point() +
  geom_line(aes(y = fitted, group = Subjects)) +
  labs(title = "Fitted Values for Each Subject", x = "Weeks", y = "Weight")

plot_model(M2,type ="pred", terms = c("Weeks"))
```



# Model 3: Weight ~ Weeks+(0+Weeks|Subjects)

```{r model3}
M3 <- lmer(Weight ~ Weeks+(0+Weeks|Subjects),REML = F, data = df)
summary(M3)
random.effects(M3)  # ranefe(M3)
```

# Plots Model M3: Weight ~ Weeks+(0+Weeks|Subjects)

```{r M3_plots}
# Residuals vs Fitted
plot(M3, which = 1)

# Extract fitted values
df$fitted <- fitted(M3)

# Plot fitted

ggplot(df, aes(x = Weeks, y = Weight, color = Subjects)) +
  geom_point() +
  geom_line(aes(y = fitted, group = Subjects)) +
  labs(title = "Fitted Values for Each Subject", x = "Weeks", y = "Weight")

plot_model(M3,type ="pred", terms = c("Weeks"))
```


# Model 4: Weight ~ Weeks+Diet+(1|Subjects)

```{r model4}
M4 <- lmer(Weight ~ Weeks+Diet+(1|Subjects),REML = F, data = df)
summary(M4)
random.effects(M4)  # ranefe(M3)
coef(M4)
fixef(M4)
```

# Plots Model M4: Weight ~ Weeks+Diet+(1|Subjects)

```{r M4_plots}
# Residuals vs Fitted
plot(M4, which = 1)

# Extract fitted values
df$fitted <- fitted(M4)

# Plot fitted by Subjects
p <- ggplot()
p <- p + geom_point(data=df, aes(x = Weeks, y = Weight, color = Subjects))
p

plot_fit_by_subjects <- geom_line(data = df, aes(x = Weeks, y = fitted, group = Subjects)) 
q <- p + plot_fit_by_subjects
q + labs(title = "Fitted Values for Each Subject", x = "Weeks", y = "Weight")

# Plot fitted by Diet
df$fit_by_Diet <- ifelse(df$Diet=='A',fixef(M4)["(Intercept)"] + df$Weeks*fixef(M4)["Weeks"],
                         (fixef(M4)["(Intercept)"]+fixef(M4)["DietB"]) + df$Weeks*fixef(M4)["Weeks"]
                        )

plot_fit_by_diets <- geom_line(data = df, aes(x = Weeks, y = fit_by_Diet, group = Diet)) 
r <- p + plot_fit_by_diets
r + labs(title = "Fitted Values for Each Diet", x = "Weeks", y = "Weight")

plot_model(M4,type ="pred", terms = c("Weeks","Diet"))

```
# Model 5: Weight ~ Weeks*Diet+(1|Subjects)

```{r model5}
M5 <- lmer(Weight ~ Weeks*Diet+(1|Subjects),REML = F, data = df)
summary(M5)
random.effects(M5)  # ranefe(M5)
coef(M5)
fixef(M5)
```

# Plots Model M5: Weight ~ Weeks*Diet+(1|Subjects)

```{r M5_plots}
# Residuals vs Fitted
plot(M5, which = 1)

# Extract fitted values
df$fitted <- fitted(M5)

# Plot fitted by Subjects
p <- ggplot()
p <- p + geom_point(data=df, aes(x = Weeks, y = Weight, color = Subjects))
p

plot_fit_by_subjects <- geom_line(data = df, aes(x = Weeks, y = fitted, group = Subjects)) 
q <- p + plot_fit_by_subjects
q + labs(title = "Fitted Values for Each Subject", x = "Weeks", y = "Weight")

# Plot fitted by Diet
df$fit_by_Diet <- ifelse(df$Diet=='A',fixef(M5)["(Intercept)"] + df$Weeks*fixef(M5)["Weeks"],
                         (fixef(M5)["(Intercept)"]+fixef(M5)["DietB"]) +
                           df$Weeks*(fixef(M5)["Weeks"]+fixef(M5)["Weeks:DietB"])
                        )

plot_fit_by_diets <- geom_line(data = df, aes(x = Weeks, y = fit_by_Diet, group = Diet)) 
r <- p + plot_fit_by_diets
r + labs(title = "Fitted Values for Each Diet", x = "Weeks", y = "Weight")

plot_model(M5,type ="pred", terms = c("Weeks","Diet"))

```
# Model 6: Weeks+Diet+(1|Subjects)

```{r model6}
M6 <- lmer(Weight ~ Weeks+Diet+(1|Subjects),REML = F, data = df)
summary(M6)
random.effects(M6)  # ranefe(M6)
coef(M6)
fixef(M6)
```

# Plots Model M6: Weeks+Diet+(1|Subjects)

```{r M6_plots}
# Residuals vs Fitted
plot(M6, which = 1)

# Extract fitted values
df$fitted <- fitted(M6)

# Plot fitted by Subjects
p <- ggplot()
p <- p + geom_point(data=df, aes(x = Weeks, y = Weight, color = Subjects))
p

plot_fit_by_subjects <- geom_line(data = df, aes(x = Weeks, y = fitted, group = Subjects)) 
q <- p + plot_fit_by_subjects
q + labs(title = "Fitted Values for Each Subject", x = "Weeks", y = "Weight")

# Plot fitted by Diet
df$fit_by_Diet <- ifelse(df$Diet=='A',fixef(M6)["(Intercept)"] + df$Weeks*fixef(M6)["Weeks"],
                         (fixef(M6)["(Intercept)"]+fixef(M6)["DietB"]) +
                           df$Weeks*fixef(M6)["Weeks"]
                        )

plot_fit_by_diets <- geom_line(data = df, aes(x = Weeks, y = fit_by_Diet, group = Diet)) 
r <- p + plot_fit_by_diets
r + labs(title = "Fitted Values for Each Diet", x = "Weeks", y = "Weight")

# Plot fitted by Diet
plot_model(M6,type ="pred", terms = c("Weeks","Diet"))

```

# Model 7: Weeks*Diet+(1|Subjects)

```{r model7}
M7 <- lmer(Weight ~ Weeks*Diet+(1|Subjects),REML = F, data = df)
summary(M7)
random.effects(M7)  # ranefe(M7)
coef(M7)
fixef(M7)
```

# Plots Model M7: Weeks*Diet+(1|Subjects)

```{r M7_plots}
# Residuals vs Fitted
plot(M7, which = 1)

# Extract fitted values
df$fitted <- fitted(M7)

# Plot fitted by Subjects
p <- ggplot()
p <- p + geom_point(data=df, aes(x = Weeks, y = Weight, color = Subjects))
p

plot_fit_by_subjects <- geom_line(data = df, aes(x = Weeks, y = fitted, group = Subjects)) 
q <- p + plot_fit_by_subjects
q + labs(title = "Fitted Values for Each Subject", x = "Weeks", y = "Weight")

# Plot fitted by Diet
df$fit_by_Diet <- ifelse(df$Diet=='A',fixef(M7)["(Intercept)"] + df$Weeks*fixef(M7)["Weeks"],
                         (fixef(M7)["(Intercept)"]+fixef(M7)["DietB"]) +
                           df$Weeks*(fixef(M7)["Weeks"]+fixef(M7)["Weeks:DietB"])
                        )

plot_fit_by_diets <- geom_line(data = df, aes(x = Weeks, y = fit_by_Diet, group = Diet)) 
r <- p + plot_fit_by_diets
r + labs(title = "Fitted Values for Each Diet", x = "Weeks", y = "Weight")

# Plot fitted by Diet
plot_model(M7,type ="pred", terms = c("Weeks","Diet"))

```
# Anova M6 vs M7: 
```{r anovaM6M7}
anova(M6,M7)
```


# Model 8: Weeks+Diet+(1+Weeks|Subjects)

```{r model8}
M8 <- lmer(Weight ~ Weeks+Diet+(1+Weeks|Subjects),REML = F, data = df)
summary(M8)
random.effects(M8)  # ranefe(M8)
coef(M8)
fixef(M8)
```

# Plots Model M8: Weeks+Diet+(1+Weeks|Subjects)

```{r M8_plots}
# Residuals vs Fitted
plot(M8, which = 1)

# Extract fitted values
df$fitted <- fitted(M8)

# Plot fitted by Subjects
p <- ggplot()
p <- p + geom_point(data=df, aes(x = Weeks, y = Weight, color = Subjects))
p

plot_fit_by_subjects <- geom_line(data = df, aes(x = Weeks, y = fitted, group = Subjects)) 
q <- p + plot_fit_by_subjects
q + labs(title = "Fitted Values for Each Subject", x = "Weeks", y = "Weight")

# Plot fitted by Diet
plot_model(M8,type ="pred", terms = c("Weeks","Diet"))

```

# Model 9: Weeks*Diet+(1+Weeks|Subjects)

```{r model9}
M9 <- lmer(Weight ~ Weeks*Diet+(1+Weeks|Subjects),REML = F, data = df)
summary(M9)
random.effects(M9)  # ranefe(M9)
coef(M9)
fixef(M9)
```

# Plots Model M9: Weeks*Diet+(1+Weeks|Subjects)

```{r M9_plots}
# Residuals vs Fitted
plot(M9, which = 1)

# Extract fitted values
df$fitted <- fitted(M9)

# Plot fitted by Subjects
p <- ggplot()
p <- p + geom_point(data=df, aes(x = Weeks, y = Weight, color = Subjects))
p

plot_fit_by_subjects <- geom_line(data = df, aes(x = Weeks, y = fitted, group = Subjects)) 
q <- p + plot_fit_by_subjects
q + labs(title = "Fitted Values for Each Subject", x = "Weeks", y = "Weight")

# Plot fitted by Diet
plot_model(M9,type ="pred", terms = c("Weeks","Diet"))

```





















