---
title: "Basic Example"
author: "Ruben Cabrera"
date: "2024-07-24"
output: 
  github_document: default
  html_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all") })
# always_allow_html: true
# always_allow_html: true
# devtools::install_github("kupietz/kableExtra")
# install.packages("xfun", repos=c("http://rstudio.org/_packages", "http://cran.rstudio.com"))
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic Example

Basic Example

```{r, message=FALSE, warning=FALSE}
library(nlmixr2)
library(ggplot2)
library(dplyr) # For 
```

## Data

Multiple dose theophylline PK data

This data set is the day 1 concentrations of the theophylline data.
A data frame with 144 rows by 7 columns

- ID: Subject ID
- TIME: Time (hrs)
- DV: Dependent Variable, theophylline concentration
- AMT: Dose Amount/kg
- EVID: rxode2/nlmixr2 event ID (not NONMEM's)
- CMT: nCompartment Number
- WT: Weight (kg)

```{r data}
?theo_sd
head(theo_sd)
summary(theo_sd)
str(theo_sd)
```


## Plot data

```{r plotdata}
ggplot(subset(theo_sd, EVID==0), aes(x = TIME, y = DV, group = factor(ID), color = factor(ID))) + geom_line() +  ylim(c(0, 12)) + xlab('Time')

ggplot(filter(theo_sd,EVID==0), aes(x=TIME, y = DV, group=factor(ID), color=factor(ID))) + geom_line() + geom_point() + ylim(c(0, 12)) + xlab('Time')
```

## Model

```{r initial}

one.compartment <- function() {

#The basic model consists of an ini block that has initial estimates  
ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv <- log(31.5); label("V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  
# and a model block with the error specification and model specification
model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    d/dt(depot) <- -ka * depot
    d/dt(center) <- ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}

```


The fit is performed by the function nlmixr/nlmixr2 specifying the model, data and estimate

```{r fit}
fit <- nlmixr2(one.compartment, theo_sd,  est="saem", saemControl(print=0))
print(fit)

```

Output description

```{r output}

# Estimation of DV: PRED and IPRED

mydata <- theo_sd[theo_sd$EVID==0,]
plot(mydata$DV,fit$PRED)
plot(mydata$DV,fit$IPRED)

# What is the difference between PRED and IPRED?
  
# 
plot(mydata$TIME,fit$DV)
plot(mydata$ID,fit$eta.ka)
plot(mydata$ID,fit$eta.cl)
plot(mydata$ID,fit$eta.v)
plot(mydata$ID,fit$cp)
plot(mydata$DV,fit$cp)
plot(mydata$DV,fit$depot)
plot(mydata$TIME,fit$depot)
plot(mydata$DV,fit$center) 
plot(mydata$TIME,fit$center)
plot(mydata$DV,fit$ka)
plot(mydata$DV,fit$cl)
plot(mydata$DV,fit$v)
plot(fit$tad,mydata$DV)
plot(mydata$DV,fit$depot)

```


Output description

```{r outputfit}

# fit
plot(fit)
```



Plots

```{r plot_model}
library(xpose.nlmixr2)
xpdb = xpose_data_nlmixr(fit)


library(xpose)
plt <- dv_vs_ipred(xpdb)

library(ggPMX)
ctr = pmx_nlmixr(fit)
pmx_plot_dv_ipred(ctr)
```



