---
title: "Basic Example"
author: "Ruben Cabrera"
date: "2024-07-29"
output: 
  github_document: default
  html_document: default
# always_allow_html: true
# devtools::install_github("kupietz/kableExtra")
# install.packages("xfun", repos=c("http://rstudio.org/_packages", "http://cran.rstudio.com"))
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basic Example

Adding Covariances between random effects
https://nlmixr2.org/articles/addingCovariances.html

```{r, message=FALSE, warning=FALSE}
library(nlmixr2)
library(ggplot2)
library(ggforce) # facet_wrap_paginate
library(dplyr) # For filter 
```

## Data

Phenobarbitol data:

This is from a PK study in neonatal infants. They received multiple doses of phenobarbital for seizure prevention.
A data frame with 744 rows and 8 columns

- ID: Infant  ID
- TIME: Time of (hr)
- AMT: Dose in (ug/kg)
- WT: Weight (kg)
- APGR: A 5-minute Apgar score to measure infant health
- DV: The concentration of phenobarbitol in the serum (ug/mL)
- MDV: If the dependent variable (DV) is missing; 0 for observations, 1 for doses
- EVID: Event ID


```{r data}
# ?pheno_sd
head(pheno_sd)
summary(pheno_sd)
str(pheno_sd)
```
```{r, fig.height=5}

# Plot data AMT Dose in (ug/kg)
ggplot(filter(pheno_sd,EVID==1), aes(x=TIME, y = AMT, group=factor(ID), color=factor(ID))) + geom_line() + geom_point() + ylim(c(0, 70)) + xlab('Time')

## Plot data DV concentration of phenobarbitol

ggplot(filter(pheno_sd,EVID==0), aes(x=TIME, y = DV, group=factor(ID), color=factor(ID))) + geom_line() + geom_point() + ylim(c(0, 70)) + xlab('Time')

# Could be interesting plot AMT vs DV (using the last data recorded in both axis)

```


```{r, fig.height=10}

# The superposition of doses as vertical black lines on concentration profiles

pAMT <- geom_segment(data=filter(pheno_sd,EVID==1), aes(x=TIME, xend=TIME, y=0, yend=AMT), size=1)

pDV <- geom_point(data=filter(pheno_sd,EVID==0), aes(x=TIME, y=DV,  color=factor(ID)), size=3)

ggplot() + pAMT
ggplot() + pDV

ggplot() + pAMT + pDV + facet_wrap_paginate(ID~., ncol=3, nrow=3, page=1) + theme(legend.position="none") + ylab("Dose and Concentration (mg/L)")

```

# Model

## Compartmental Models

Pharmacokinetics is the study of the time course of a drug and its metabolites following introduction into the body. In this section we describe a class of models that has been extensively used in such studies to model individual drug concentrations, $y(t)$, as a function of the time $t$.

Let $A_i(t)$ represent the amount of drug in compartment $i$, $i=0,1$ at time $t$, with compartment 0 representing the site from which absorption occurs. The differential equations describing the drug flow between the compartments may be assumed to be of the form

$$ \frac{d A_0}{d t} = -k_a A_0
$$

$$ \frac{d A_1}{d t} = k_a A_0 - k_e A_1
$$

where $k_a$ is the absorption rate constant associated with flow from compartment 0 to compartment 1, and $k_e$ is the elimination rate constant.

With only one compartment $A_1$, the differential equation for the amount $A_1$ is

$$ \frac{d A_1}{dt } = - k_e A_1 $$

The concentration is 

$$C_p = \frac{A_1(t)}{V(t)}$$
and considering $V$ as volume, the clearance $ CL $ is defined from

$$k_e = \frac{CL}{V}$$

It is usual to parametrize the model in terms of clearance $CL$ and volume $V$ instead of $k_e$.

The actual fitting parametrization is carried out in term of so called typical values for clearance and volume $tcl$ and $tv$ :


$$
\begin{align}
  CL =& e^{ tcl + \eta_{CL} }  \\
  V  =& e^{ tv  + \eta_{V} } 
\end{align}
$$
where $\eta_{CL}$ and $\eta_{V}$ are random variables obeying normal distributions with mean zero and covariance matrix $\Omega$. The components of $\Omega$ are then additional fit parameters. The matrix $\Omega$ and the variables $\eta$ are responsible to explain random inter-individual variability.

$$(\eta_{CL} , \eta_{V}) \approx   N(0, \Omega)  $$
If $\Omega$ is a diagonal matrix, the inter-individual variabilities are independent. Otherwise, they have some correlation.

The residual error model provides the final random variability on the predicted concentration. The residual error model in this example is

$$C_{obs} = C_p + \varepsilon$$

where $C_{obs}$ is the observed concentration, $C_p$  is the concentration in plasma and $\varepsilon$ is a normal random variable (specified by $add.err=\sigma^2$).

$$ \varepsilon \approx N(0, \sigma^2) $$

```{r initial}
pheno <- function() {
  ini({
    tcl <- log(0.008) # typical value of clearance
    tv <-  log(0.6)   # typical value of volume
    eta.cl + eta.v ~ c(1, 0.01, 1) ## var(eta.cl), cov(eta.cl, eta.v), var(eta.v)
                      # interindividual variability on clearance and volume
    add.err <- 0.1    # residual variability
  })
  model({
    cl <- exp(tcl + eta.cl) # individual value of clearance
    v <- exp(tv + eta.v)    # individual value of volume
    ke <- cl / v            # elimination rate constant
    d/dt(A1) = - ke * A1    # model differential equation
    cp = A1 / v             # concentration in plasma
    cp ~ add(add.err)       # define error model
  })
}

```


The fit is performed by the function nlmixr/nlmixr2 specifying the model, data and estimate

```{r fit}
fit <- nlmixr(pheno, pheno_sd, "saem", control=list(print=0), table=list(cwres=TRUE, npde=TRUE))
print(fit)
str(fit)

```


# Output description


```{r output}

# Estimation of DV: PRED and IPRED

mydata <- pheno_sd[pheno_sd$EVID==0,]
plot(mydata$DV,fit$PRED)
plot(mydata$DV,fit$IPRED)
plot(mydata$DV,fit$EPRED)

# What is the difference between PRED and IPRED?
  
# eta.ka:
plot(mydata$ID,fit$eta.ka)
plot(mydata$ID,fit$eta.cl)
plot(mydata$ID,fit$eta.v)
plot(mydata$ID,fit$cp)
plot(mydata$DV,fit$cp)
plot(mydata$DV,fit$depot)
plot(mydata$DV,fit$center)   
plot(mydata$DV,fit$ka)
plot(mydata$DV,fit$cl)
plot(mydata$DV,fit$v)
plot(mydata$DV,fit$tad)
plot(mydata$DV,fit$depot)
plot(mydata$DV,fit$dosenum)
```



Plots

```{r plot_model}
library(xpose.nlmixr2)
xpdb = xpose_data_nlmixr(fit)


library(xpose)
plt <- dv_vs_ipred(xpdb)

library(ggPMX)
ctr = pmx_nlmixr(fit)
pmx_plot_dv_ipred(ctr)
```

Plots

```{r plot_fit}
plot(fit)
```

```{r}
plot(augPred(fit))
```
```{r}
p1 <- vpcPlot(fit, show=list(obs_dv=TRUE));
#> [====|====|====|====|====|====|====|====|====|====] 0:00:00
p1 <- p1 + ylab("Concentrations")

## A prediction-corrected VPC
p2 <- vpcPlot(fit, pred_corr = TRUE, show=list(obs_dv=TRUE))
#> [====|====|====|====|====|====|====|====|====|====] 0:00:00
p2 <- p2 + ylab("Prediction-Corrected Concentrations")

library(patchwork)
p1 / p2
```


# Modified Model

We change the fitting parametrization with typical values for clearance and volume $tcl$ and $tv$ and also a new variable $WT_{CL}$


$$
\begin{align}
  CL =& e^{ tcl + \eta_{CL} } NWE^{WT_{CL}}  \\
  V  =& e^{ tv  + \eta_{V} } 
\end{align}
$$
Where $NWT$ is the weights normalized bye mean $WT$ so:

```{r }

# Calculate median
a <- median(distinct(select(pheno_sd,ID,WT))$WT)
pheno_sd$NWE <- pheno_sd$WT - a
```


```{r}
m2 <- function() {
  ini({
    tcl <- log(0.008) # typical value of clearance
    tv <-  log(0.6)   # typical value of volume
    wt.cl <- 0.75
    eta.cl + eta.v ~ c(1, 0.01, 1) ## var(eta.cl), cov(eta.cl, eta.v), var(eta.v)
                      # interindividual variability on clearance and volume
    add.err <- 0.1    # residual variability
  })
  model({
    cl <- exp(tcl + eta.cl)*NWE^wt.cl  
    v <- exp(tv + eta.v)     # individual value of volume
    ke <- cl / v             # elimination rate constant
    d/dt(A1) = - ke * A1     # model differential equation
    cp = A1 / v              # concentration in plasma
    cp ~ add(add.err)        # define error model
  })
}

```

The fit is performed by the function nlmixr/nlmixr2 specifying the model, data and estimate

```{r}
##fit2 <- nlmixr(m2, pheno_sd, "saem", control=list(print=0), table=list(cwres=TRUE, npde=TRUE))
##print(fit2)
##str(fit2)

```