---
title: "NLMIXR_Intro"
author: "Renan Cabrera"
date: "2023-07-01"
output: 
  github_document: default
  html_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
rm(list = ls())

library(dplyr)
library(ggplot2)
library(ggforce)

library(ggstance)

library(tidyr)
library(comprehenr)

library(kableExtra)

library(purrr)

library(nlmixr2)
```

# Introduction
This documents is an introduction of Non Linear Mixed Effects with the help of the R package NLMIXR2.
This example is based on: 
https://nlmixr2.org/articles/addingCovariances.html

# Data
The data corresponds to Phenobarbitol, which is automatically loaded with MLMIXR 
```{r}
#pheno_sd
str(pheno_sd)
```
The columns are defined as

* ID: Identification of individual. 
* TIME: Time 
* AMT:  Amount of drug dose.
* WT:   Weight.  This is a covariate
* APGR: Apgar score.  This is a covariate
* DV:   Dependent variable observed. This is the measured concentration in blood plasma.
* MDV:  Missing dependent variable. This is 1 whenever the row value is not an DV observation.
        If MDV=0, DV must be used, otherwise DV must be ignored. 
* EVID: EVID=0 for dependent variable observed. 
        EVID=1 for dose record. The AMT column should be used.
        
The naming of the columns ID, TIME, AMT, DV, MDV and EVID follows the NONMEN convention. 
More information can be found at: 

* Bauer, Robert J. "NONMEM tutorial part I: description of commands and options, with simple examples of population analysis." CPT: pharmacometrics & systems pharmacology 8.8 (2019): 525-537.


```{r}
p <- ggplot( data= filter(pheno_sd,EVID==0),  
             aes(x=TIME,y=DV, group=factor(ID), color=factor(ID) ))
p <- p + geom_line() + geom_point()
p <- p + theme_bw()
p <- p + xlab("Time (h)") + ylab("Concentration (mg/L)")
p #+ facet_wrap(WT~.)
```
The superposition of doses as vertical black lines on concentration profiles
```{r, fig.height=10}
p <- ggplot( )
p <- p +  geom_segment( 
  data= filter(pheno_sd,EVID==1),
  aes(x=TIME, xend=TIME, y=0,yend=AMT), size=1 )

p <- p +  geom_point(
  data= filter(pheno_sd,EVID==0),
  aes(x=TIME, y=DV, color=factor(ID) ), size=3 )

p <- p + theme_bw()
p <- p + xlab("Time (h)") + ylab("Concentration (mg/L)")
p <- p + theme(legend.position="none")
p + facet_wrap_paginate(ID~., ncol=3, nrow=3, page=1)
```


# Model 

The differential equation for the amount is
$$ \frac{A_1}{dt } = - k_e A_1 $$
The concentration is 
$$C_p = \frac{A_1}{V}$$

The clearance is defined from 
$$k_e = \frac{CL}{V}$$
It is usual to parametrize the model in terms of clearance and volume instead of $k_e$.

The actual fitting parametrization is carried out in term of so called typical values $LogTVCL$ and $LogTVV$:

$$
\begin{align}
  CL =& \exp( LogTVCL + \eta_{CL} )  \\
  V  =& \exp( LogTVV  + \eta_{V} ) 
\end{align}
$$
where $\eta_{CL}$ and $\eta_{V}$ are random variables obeying normal distributions with a covariance matrix $\Omega$ (mean zero). The components of $\Omega$ are then additional fit parameters. The matrix $\Omega$ and the variables $\eta$ are responsible to explain random inter-individual variability.

$$(\eta_{CL} , \eta_{V}) \approx   N(0, \Omega)  $$
If $\Omega$ is a diagonal matrix, the inter-individual variabilities are independent. Otherwise, they have some correlation.



The residual error model provides the final random variability on the predicted concentration. The residual error model
in this example is 
$$C_{obs} = C_p + \varepsilon$$,
where $\varepsilon$ is a normal random variable (specified by $add.err=\sigma^2$).
$$ \varepsilon \approx N(0, \sigma^2) $$

```{r}
pheno <- function() {
  ini({
    LogTVCL <- log(0.008) # typical value of log clearance with initial value CL=0.008
    LogTVV <-  log(0.6)   # typical value of log volume with initial value V=0.6 
    ## var(eta.cl)
    eta.cl + eta.v ~ c(1, 
                       0.01, 1) ## cov(eta.cl, eta.v), var(eta.v)
                      # interindividual variability on clearance and volume
    add.err <- 0.1    # residual variability
  })
  model({
    cl <- exp(LogTVCL + eta.cl) # individual value of clearance
    v <- exp(LogTVV + eta.v)    # individual value of volume
    ke <- cl / v            # elimination rate constant
    d/dt(A1) = - ke * A1    # model differential equation
    cp = A1 / v             # concentration in plasma
    cp ~ add(add.err)       # define residual error model
  })
}
```

Fit with the SAEM method (stochastic optimization)
```{r, message=FALSE}
fit <- nlmixr(pheno, pheno_sd, "saem",
              control=list(print=0), 
              table=list(cwres=TRUE, npde=TRUE))
```

```{r}
str(fit)
```

The final objective function and associated values
```{r}
fit$objDf         %>% kable %>% kable_styling()
```

The fitted parameters are
```{r}
fit$parFixedDf %>% kable %>% kable_styling()
```

where Back transformed is the exponential of the actual estimations in the log scale.

The estimated $\Omega$ covariance matrix is
```{r}
as.data.frame( fit$omega ) %>% kable %>% kable_styling()
```

Additionally, the output contains time series
```{r}
fit.time.series <- as.data.frame(fit)
```

This time series contains as many rows as many concentration observations. 
```{r}
nrow(fit.time.series)
```

```{r}
str(fit.time.series)
```
The definition of some columns:

* PRED: The concentration without inter-individual variability $\eta=0$ also known as population concentration prediction.

* IPRED: The concentration with inter-individual variability.



The superposition of doses as vertical black lines on concentration profiles
```{r, fig.height=10,fig.width=8}
p <- ggplot( )
p <- p +  geom_segment( 
  data= filter(pheno_sd,EVID==1),
  aes(x=TIME, xend=TIME, y=0,yend=AMT), size=1 )

p <- p +  geom_point(
  data= filter(pheno_sd,EVID==0),
  aes(x=TIME, y=DV ), size=3, color="#7CAE00" )

p <- p + geom_line(
    data=fit.time.series,
    aes(x=TIME,y=PRED  ), color="blue", size=1 )

p <- p + geom_line(
    data=fit.time.series,
    aes(x=TIME,y=IPRED  ), color="red" )


p <- p + theme_bw()
p <- p + xlab("Time (h)") + ylab("Concentration (mg/L)")
p <- p + theme(legend.position="none")
p <- p + labs(caption = "Red: Individual predictions \n Blue: Population predictions") 
p + facet_wrap_paginate(ID~., ncol=3, nrow=3, page=1)
```

A time series with higher resolution in time can be calculated as

```{r}
fit.time.series.aug <- mutate( as.data.frame( augPred( fit )  ), ID=id )%>% select(-id)
nrow(fit.time.series.aug)
```
```{r}
str(fit.time.series.aug)
select(fit.time.series.aug,ind)%>% distinct()
#select(fit.time.series.aug,Endpoint)%>% distinct()

```


```{r}
#plot(augPred(fit))
```

```{r, fig.height=10,fig.width=8}
p <- ggplot( )

p <- p +  geom_segment( 
  data= filter(pheno_sd,EVID==1),
  aes(x=TIME, xend=TIME, y=0,yend=AMT), size=1 )

p <- p +  geom_point(
  data= filter(pheno_sd,EVID==0),
  aes(x=TIME, y=DV ), size=3, color="#7CAE00" )

p <- p + geom_line(
    data = filter( fit.time.series.aug, ind=="Individual") ,
    aes(x=time,y=values  ), color="red", size=0.5 )

p <- p + geom_line(
    data = filter( fit.time.series.aug, ind=="Population") ,
    aes(x=time,y=values  ), color="blue", size=0.5 )

#p <- p + scale_x_continuous(limits=c(0,200))

p <- p + theme_bw()
p <- p + xlab("Time (h)") + ylab("Concentration (mg/L)")
p <- p + theme(legend.position="none")
p <- p + labs(caption = "Red: Individual predictions \n Blue: Population predictions") 
p + facet_wrap_paginate(ID~., ncol=1, nrow=4, page=1, scales = "free")
```

# Bibliography

* Nguyen, T. H. T., et al. "Model evaluation of continuous data pharmacometric models: metrics and graphics. CPT Pharmacometrics Syst Pharmacol 6: 87–109." (2017).

* Bauer, Robert J. "NONMEM tutorial part I: description of commands and options, with simple examples of population analysis." CPT: pharmacometrics & systems pharmacology 8.8 (2019): 525-537.

* Wang, Yaning. "Derivation of various NONMEM estimation methods." Journal of Pharmacokinetics and pharmacodynamics 34 (2007): 575-593.


